A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-02 tex/body.tex
more consistent naming

\begin{report}
\def\thewordpaper{report}
\renewcommand\dotReportFootnote[1]{.\footnote{#1}}
\end{report}

@@

Datatypes and codatatypes are an integral part of modern proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have made their appearance in a few automatic theorem provers. The SMT-LIB~2
\cite{barrett-et-al-2010} syntax, implemented by most SMT (satisfiability
modulo theories) solvers, includes a theory of datatypes.

@@

\begin{paper}
Proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{paper}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
%\textbf{TODO: CHECK WITH IMPLEMENTATION, but round to nearest 50 or 1000.}
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\paragraph{Conventions.}
The setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool}\begin{report} interpreted as the set of truth
values $\{\bot, \top\}$\end{report}. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$)\begin{report} and belongs to the logical symbols\end{report}.
Other predicates can be represented as functions to $\ty{bool}$,
with $\const{p}(\ldots)$ abbreviating $\const{p}(\ldots) \teq \const{true}$.
\begin{report}
The metavariables $\delta,\:\varepsilon$ range over (co)datatypes,
whereas $\tau,\:\upsilon$ range over arbitrary types.
When applied to terms, the symbol $=$ denotes syntactic equality.\end{report}

\begin{report}
Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
\end{report}
Symbols starting with an uppercase letter (e.g.,
$\const{C}, \const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed %compactly
as $\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$\begin{report},
which collapses to $\const{f} : \tau$ if $n = 0$\end{report}.
For a term $t$, the notation $t\typ{\tau}$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
\begin{report}%
The operator $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$. \end{report}%
Finally, $[n]$ denotes the set $\{1,\ldots, n\}$, and
$\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.

@@

\begin{report}%
Because all types must be inhabited (nonempty), a datatype specification is
admissible only if a ground constructor term can be exhibited.
This rules out non-well-founded specifications such as
\[\keyw{datatype}\;\, \ty{fstream}_{\,\tau} \,=\, \const{FSCons}(\tau,\:\, \ty{fstream}_{\,\tau})\]
For codatatypes, no admissibility check is necessary because there is always a term,
finite or infinite, that witnesses nonemptiness \cite{blanchette-et-al-2015-esop}.
\end{report}

A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is mutually (co)recursive if and only if the associated dependency graph
is strongly connected. Types can be declared together as mutually
(co)recursive even if they are not actually (co)recursive, but the
semantic notion is more precise and is the one that interests us.
%
In addition, nothing forbids non(co)recursive specifications
such~as
\begin{paper}%
\vthinspace$\keyw{datatype}~\vthinspace\ty{option}_{\,\tau} = \const{None} \mid \const{Some}(\tau)$.%
\end{paper}%
\begin{report}%
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{option}_{\,\tau} & \,=\, \const{None} \mid \const{Some}(\tau) \\[-.5\jot]
      \keyw{codatatype}\; \,\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
At the semantic level, it makes no difference whether such types are
introduced as datatypes or as codatatypes.
\end{report}%
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

@@

%\paragraph{Characterization.}
One way to define datatypes semantically is as the initial model of the
selector--constructor equations \cite{barrett-et-al-2010}.
\begin{report}
A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
this gives rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$. This flaw
could be corrected, but the added complexity seems to suggest that selectors
are better characterized axiomatically.

\end{report}
A related semantic view of datatypes is as initial algebras. Codatatypes are
then defined dually as final coalgebras \cite{rutten-2000}. The datatypes are
generated by their constructors, whereas the codatatypes are viewed through
their selectors.
%By uniformly focusing on the constructors, the
%axiomatic approach emphasizes the commonality between datatypes and
%codatatypes, while sacrificing a theoretically fruitful notion of duality.

Datatypes and codatatypes share many basic properties pertaining to
constructors and selectors. All properties below are implicitly universally
quantified and range over all $i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-.5\jot]
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y}
  \\[-.5\jot]
\text{Exhaustiveness:}\quad
  & \smash{\is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)}
  \\[-.5\jot]
\text{Selection:}\quad
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k}
\end{aligned}
\]
%
\begin{report}%
Expressed in the algebraic jargon, exhaustiveness helps ensure that ``no
junk'' exists, whereas distinctness and injectivity guarantee that ``no
confusion'' can arise.
The result of selectors applied to the wrong
constructor is left completely unspecified.
\end{report}

@@

Datatypes are additionally characterized by an induction axiom schema\begin{report}:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k}\, \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[\const C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$\end{report}.
%
The axiom schema ensures that the interpretation of datatypes
is a standard model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.
%
\begin{report}\par\end{report}
%
For codatatypes, the dual notion is called coinduction. \begin{report}It depends on
witnesses $R_i$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\kern.8em %%% TYPESETTING: should be \quad
\AXC{\strut$\begin{gathered}\textstyle \bigwedge_{\,i}\, R_i[v_i, w_i] \\[-\jot]\textstyle
\bigwedge_{\,i}\, \forall v\; w.\;\, R_i[v, w] \longrightarrow
  \bigwedge_{\,j}\, \const{d}_{\negvthinspace j}(v) \teq \const{d}_{\negvthinspace j}(w)
  \mathrel\land
  \const{d}_{\negvthinspace j}(v) \longrightarrow \bigwedge_{\,k}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\,i}\, v_i \teq w_i$}
\DP
\end{aligned}
\]
where $x \sim y$ denotes either $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
\end{report}%
This axiom schema guarantees that two values that yield the same
observations must be equal, where the observations are made by using the
selectors and discriminators.
%
\begin{report}\par\end{report}
  %
In addition, codatatypes are guaranteed to contain all values corresponding to
infinite ground constructor terms.
\begin{report}
In general, this cannot be captured by a
first-order axiomatization, since there may be uncountably many of them.
For example, $\ty{stream}_{\,\ty{int}}$ is isomorphic to the uncountable
function space $\ty{nat} \to \ty{int}$.
\end{report}

@@

\begin{report}
TODO: State more precisely and proof?
\end{report}

@@

%\paragraph{Special Cases.}
Some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\vthinspace$\keyw{codatatype}~\ty{a} = \const{A}(\ty{a})$, whose
only value is $\MU a.\; A(a)$. Other specimens are
\begin{paper}%
\vthinspace$\keyw{codatatype}~\ty{b} = \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit})
~\keyw{and}~ \ty{c} = \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})$,
\end{paper}%
\begin{report}%
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-1\jot]
      \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
\end{report}%
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types. %, even if they rarely arise in practice.
A type may also be a corecursive singleton only in some models. If the example
above is altered to leave \ty{unit} uninterpreted, \ty{b} and \ty{c} will be
singletons precisely when \ty{unit} is interpreted as a singleton.
Fortunately, given cardinalities for the ordinary types, it is easy to
characterize this degenerate case:

@@

\begin{report}
\begin{proof}
By definition, the type is equipped with at least one (directly or indirectly)
corecursive constructor $\const{C}$. If it additional has second
corecursive constructor $\const{D}$, it is possible to encode infinitely many
alternation patterns---e.g.,
$\const{C}(\const{D}(\const{C}(\const{C}(\ldots))))$---all of which correspond
to distinct values (by distinctness and injectivity). If the type has a
noncorecursive constructor $\const{E}$, it is possible to create terms of
arbitrary depths---e.g., $\const{C}(\ldots(\const{C}(\const{E}))\ldots)$. In
both cases, there can be no finite models.

Therefore, $\const{C}$ must be the only constructor.
If any of its noncorecursive arguments has a cardinality greater than 1,
it is possible to encode alternation patterns using it---e.g.,
$\const{C}(0,\: \const{C}(1,\: \const{C}(0,\: \const{C}(0,\: \ldots))))$---which
again excludes finite models. Otherwise, the coinduction principle ensures
that the type has at most one value.
\qed
\end{proof}
\end{report}

@@

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$\begin{report},
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{report}.
Two $\mu$-terms can represent the same value while not being $\alpha$-equivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{C}( \const{C}( x ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

\begin{report}
\begin{lemma}
\label{lem:mu-norm-arg}
If the $\mu$-terms $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim \interp{t}{\vec w}$\vthinspace.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

%\begin{lemma}
%\label{lem:mu-cong}
%If $t_u = \MU x.\; \const{C}( \vec u )$ and $t_w = \MU y.\; \const{C}( \vec w )$ are in normal form,
%then $\interp{t_u}{ \vec u } \vsim \interp{t_w}{ \vec w }$ if and only if $t_u \vsim t_w$.
%\end{lemma}
%\begin{proof}
%\rem{TODO}
%\end{proof}
\end{report}

@@

%%% @ANDY: I think it's enough to point this out clearly in the introduction.
%%% At least, that always worked fine for my other papers. :)
%
%\begin{paper} 
%The proof of this lemma can be found in the extended version of this paper.
%\end{paper}

>>>>

\begin{rep}
\def\thewordpaper{report}
\renewcommand\dotReportFootnote[1]{.\footnote{#1}}
\end{rep}

@@

Datatypes and codatatypes are an integral part of most modern proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have made their appearance in a few automatic theorem provers. The SMT-LIB~2
\cite{barrett-et-al-2010} syntax, implemented by most SMT (satisfiability
modulo theories) solvers, includes a theory of datatypes.

@@

\begin{conf}
Proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{conf}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
%\textbf{TODO: CHECK WITH IMPLEMENTATION, but round to nearest 50 or 1000.}
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

\paragraph{Conventions.}
The setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool}\begin{rep} interpreted as the set of truth
values $\{\bot, \top\}$\end{rep}. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$)\begin{rep} and belongs to the logical symbols\end{rep}.
Other predicates can be represented as functions to $\ty{bool}$,
with $\const{p}(\ldots)$ abbreviating $\const{p}(\ldots) \teq \const{true}$.
\begin{rep}
The metavariables $\delta,\:\varepsilon$ range over (co)datatypes,
whereas $\tau,\:\upsilon$ range over arbitrary types.
When applied to terms, the symbol $=$ denotes syntactic equality.\end{rep}

\begin{rep}
Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
\end{rep}
Symbols starting with an uppercase letter (e.g.,
$\const{C}, \const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed %compactly
as $\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$\begin{rep},
which collapses to $\const{f} : \tau$ if $n = 0$\end{rep}.
For a term $t$, the notation $t\typ{\tau}$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
\begin{rep}%
The operator $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$. \end{rep}%
Finally, $[n]$ denotes the set $\{1,\ldots, n\}$, and
$\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.

@@

\begin{rep}%
Because all types must be inhabited (nonempty), a datatype specification is
admissible only if a ground constructor term can be exhibited.
This rules out non-well-founded specifications such as
\[\keyw{datatype}\;\, \ty{fstream}_{\,\tau} \,=\, \const{FSCons}(\tau,\:\, \ty{fstream}_{\,\tau})\]
For codatatypes, no admissibility check is necessary because there is always a term,
finite or infinite, that witnesses nonemptiness \cite{blanchette-et-al-2015-esop}.
\end{rep}

A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is mutually (co)recursive if and only if the associated dependency graph
is strongly connected. Types can be declared together as mutually
(co)recursive even if they are not actually (co)recursive, but the
semantic notion is more precise and is the one that interests us.
%
In addition, nothing forbids non(co)recursive specifications
such~as
\begin{conf}%
\vthinspace$\keyw{datatype}~\vthinspace\ty{option}_{\,\tau} = \const{None} \mid \const{Some}(\tau)$.%
\end{conf}%
\begin{rep}%
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{option}_{\,\tau} & \,=\, \const{None} \mid \const{Some}(\tau) \\[-.5\jot]
      \keyw{codatatype}\; \,\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
At the semantic level, it makes no difference whether such types are
introduced as datatypes or as codatatypes.
\end{rep}%
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

@@

%\paragraph{Characterization.}
One way to define datatypes semantically is as the initial model of the
selector--constructor equations \cite{barrett-et-al-2010}.
\begin{rep}
A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
this gives rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$. This flaw
could be corrected, but the added complexity seems to suggest that selectors
are better characterized axiomatically.

\end{rep}
A related semantic view of datatypes is as initial algebras. Codatatypes are
then defined dually as final coalgebras \cite{rutten-2000}. The datatypes are
generated by their constructors, whereas the codatatypes are viewed through
their selectors.
%By uniformly focusing on the constructors, the
%axiomatic approach emphasizes the commonality between datatypes and
%codatatypes, while sacrificing a theoretically fruitful notion of duality.

Datatypes and codatatypes share many basic properties pertaining to
constructors and selectors. All properties below are implicitly universally
quantified and range over all $i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-.5\jot]
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y}
  \\[-.5\jot]
\text{Exhaustiveness:}\quad
  & \smash{\is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)}
  \\[-.5\jot]
\text{Selection:}\quad
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k}
\end{aligned}
\]
%
\begin{rep}%
Expressed in the algebraic jargon, exhaustiveness helps ensure that ``no
junk'' exists, whereas distinctness and injectivity guarantee that ``no
confusion'' can arise.
The result of selectors applied to the wrong
constructor is left completely unspecified.
\end{rep}

@@

Datatypes are additionally characterized by an induction axiom schema\begin{rep}:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k}\, \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[\const C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$\end{rep}.
%
The axiom schema ensures that the interpretation of datatypes
is a standard model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.
%
\begin{rep}\par\end{rep}
%
For codatatypes, the dual notion is called coinduction. \begin{rep}It depends on
witnesses $R_i$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\kern.8em %%% TYPESETTING: should be \quad
\AXC{\strut$\begin{gathered}\textstyle \bigwedge_{\,i}\, R_i[v_i, w_i] \\[-\jot]\textstyle
\bigwedge_{\,i}\, \forall v\; w.\;\, R_i[v, w] \longrightarrow
  \bigwedge_{\,j}\, \const{d}_{\negvthinspace j}(v) \teq \const{d}_{\negvthinspace j}(w)
  \mathrel\land
  \const{d}_{\negvthinspace j}(v) \longrightarrow \bigwedge_{\,k}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\,i}\, v_i \teq w_i$}
\DP
\end{aligned}
\]
where $x \sim y$ denotes either $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
\end{rep}%
This axiom schema guarantees that two values that yield the same
observations must be equal, where the observations are made by using the
selectors and discriminators.
%
\begin{rep}\par\end{rep}
  %
In addition, codatatypes are guaranteed to contain all values corresponding to
infinite ground constructor terms.
\begin{rep}
In general, this cannot be captured by a
first-order axiomatization, since there may be uncountably many of them.
For example, $\ty{stream}_{\,\ty{int}}$ is isomorphic to the uncountable
function space $\ty{nat} \to \ty{int}$.
\end{rep}

@@

\begin{rep}
TODO: State more precisely and proof?
\end{rep}

@@

%\paragraph{Special Cases.}
Some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\vthinspace$\keyw{codatatype}~\ty{a} = \const{A}(\ty{a})$, whose
only value is $\MU a.\; A(a)$. Other specimens are
\begin{conf}%
\vthinspace$\keyw{codatatype}~\ty{b} = \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit})
~\keyw{and}~ \ty{c} = \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})$,
\end{conf}%
\begin{rep}%
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-1\jot]
      \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
\end{rep}%
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types. %, even if they rarely arise in practice.
A type may also be a corecursive singleton only in some models. If the example
above is altered to leave \ty{unit} uninterpreted, \ty{b} and \ty{c} will be
singletons precisely when \ty{unit} is interpreted as a singleton.
Fortunately, given cardinalities for the ordinary types, it is easy to
characterize this degenerate case:

@@

\begin{rep}
\begin{proof}
By definition, the type is equipped with at least one (directly or indirectly)
corecursive constructor $\const{C}$. If it additional has second
corecursive constructor $\const{D}$, it is possible to encode infinitely many
alternation patterns---e.g.,
$\const{C}(\const{D}(\const{C}(\const{C}(\ldots))))$---all of which correspond
to distinct values (by distinctness and injectivity). If the type has a
noncorecursive constructor $\const{E}$, it is possible to create terms of
arbitrary depths---e.g., $\const{C}(\ldots(\const{C}(\const{E}))\ldots)$. In
both cases, there can be no finite models.

Therefore, $\const{C}$ must be the only constructor.
If any of its noncorecursive arguments has a cardinality greater than 1,
it is possible to encode alternation patterns using it---e.g.,
$\const{C}(0,\: \const{C}(1,\: \const{C}(0,\: \const{C}(0,\: \ldots))))$---which
again excludes finite models. Otherwise, the coinduction principle ensures
that the type has at most one value.
\qed
\end{proof}
\end{rep}

@@

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$\begin{rep},
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can represent the same value while not being $\alpha$-equivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{C}( \const{C}( x ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

\begin{rep}
\begin{lemma}
\label{lem:mu-norm-arg}
If the $\mu$-terms $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim \interp{t}{\vec w}$\vthinspace.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

%\begin{lemma}
%\label{lem:mu-cong}
%If $t_u = \MU x.\; \const{C}( \vec u )$ and $t_w = \MU y.\; \const{C}( \vec w )$ are in normal form,
%then $\interp{t_u}{ \vec u } \vsim \interp{t_w}{ \vec w }$ if and only if $t_u \vsim t_w$.
%\end{lemma}
%\begin{proof}
%\rem{TODO}
%\end{proof}
\end{rep}

@@

%%% @ANDY: I think it's enough to point this out clearly in the introduction.
%%% At least, that always worked fine for my other papers. :)
%
%\begin{conf} 
%The proof of this lemma can be found in the extended version of this paper.
%\end{conf}

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal model $\M$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm (2)]
$\ValC \ec{x} = \interp{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$. 
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{rep}
\begin{proof}
To show 1, we first show that $\ValC$ contains no free variables.
Assume by contradiction that $\ValC$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\ValC)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\ValC)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\ValC)$ to ${\Varec{y}}$.
By construction, $\ValC \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\ValC$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\ValC \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.
